(Re)building cooperation: Effects of a cognitive intervention on cooperative behavior in games

### 2024-09-07 article.Rmd
new draft version1

abstract: |
  Social trust is a key component of strong social bonds. Its absence is a risk factor for social dysfunction and psychopathology. As such, interventions to foster and strengthen trust-based cooperation are highly desirable. We used the Repeated Trust Game to evaluate a cognitive intervention aiming to repair cooperation breakdowns caused by a one-time defection by the co-player. In this study, 318 participants played the role of trustee across two online games, believing they were facing two different players. Between games, some received a brief cognitive intervention. Post-intervention, participants showed more cooperative behavior both before and after defection by the co-player. Analysis using a hidden Markov model revealed that those in the intervention group had a higher proportion of cooperative states than participants in the control group. This is consistent with participants inferring from the intervention that pro-social behavior may generally provide more beneficial outcomes to them in the long run.

@@

Cooperation refers to the collaborative efforts of individuals or entities working towards a shared goal or advantage [@tomasello2012two]. This coordinated approach often leads to outcomes that exceed the sum of individual contributions.  Trust is a key driver of cooperation, enabling individuals to engage in mutually beneficial interactions without fear of exploitation [@balliet2013trust]. This aligns with some definitions of trust as “a psychological state comprising the intention to accept vulnerability based on positive expectations of the intentions or behavior of another” [@rousseau1998not]. When we trust others, we are more likely to cooperate with them, even when there is a risk of being cheated. Prior research also emphasises the importance of social trust in determining why some people fare better than others physically and mentally [@giordano_trust_2016; @meng_multilevel_2014]. Research into the determinants of psychopathology has linked trust-based constructs to the emergence of mental health disorders [@fonagy_role_2014].

A well-established paradigm in the study of trust is the Repeated Trust Game (RTG; @joyce_trust_1995). In this game, the "investor" decides how much of an endowment to send to the other player (the "trustee"). The amount that is sent is tripled and the trustee decides, in return, how much of the tripled amount to send back to the investor. The Nash equilibrium for a single-round version of the game is for the investor to send nothing. In the repeated version, rewards for both players are maximised if they build trust and share the benefits of higher investments. If the investor is rewarded for taking the risk of sending an investment, they are likely invest again in future rounds. But if the investor obtains a low return on their investment, they will likely reduce future investments, thereby diminishing the potential gains for both players.

Since a lack of epistemic trust is a risk factor for social dysfunction [@fonagy_mentalizing_2017], and given the importance of trust for building and maintaining strong social bonds, interventions that foster and strengthen trust-based cooperation would be highly beneficial to society. Such interventions would allow people to more easily repair broken relationships and continue harvesting the benefits of cooperation even in the presence of accidental or intentional social norm violations.

To encourage the emergence and maintenance of trust in the RTG setting, some studies have focused on modification of the game, e.g. through introducing a third-party who monitors the actions of the other players [@charness_investment_2008; @fiedler_effect_2017]. Others chose to prime participants towards feeling gratitude [@drazkowski_gratitude_2017] or with the concepts of friend and foe [@burnham_friend-or-foe_2000]. Whilst these results show that it is possible to improve cooperative outcomes at the start of the game, they do not address how to repair a breakdown of trust after intentional or accidental non-cooperative actions by the players. Cooperation in the RTG can easily break down after a transgressive behavior such as a nil or very low investment by the investor or a return by the trustee below the investment sent [@bendor_when_1991]. Such breakdowns of cooperation are particularly evident when the trustee suffers from social disorders such as Borderline Personality Disorder (BPD; @lieb_borderline_2004). Trustees with BPD fail to engage in trust-repairing behaviors such as coaxing the investor by signalling trustworthiness via sending high returns [@king-casas_rupture_2008].  Neuroimaging analysis from the study of @king-casas_rupture_2008 is consistent with the idea that individuals with BPD may not recognize how offering low returns in such interactions conflicts with societal expectations. Further, computational analysis by @hula2018model links this issue to some BPD individuals' mentalizing difficulties, such as their inability to foresee the impact of their low returns on the investor's future actions.

In devising interventions to repair trust in the RTG, we can take inspiration from cognitive interventions championed by successful therapies that aim to improve social dysfunction in BPD. Psychotherapies such as Mentalisation Based Therapy [@allen_handbook_2006-1] and Dialectical Behavior Therapy [@linehan_cognitive-behavioral_1993] have been shown to improve social skills in BPD [@gunderson_borderline_2018]. However, response to these treatments is highly variable, and determining which interventions are effective for particular sub-groups of patients is challenging [@rudge_mechanisms_2020; @arch_longitudinal_2012]. One promising approach is the study of how specific components of psychotherapeutic treatment affect quantitative markers of behavior such as those inferred through computational models [@huys_computational_2016; @reiter_neuro-cognitive_2021; @dercon2022core]. Combining the use of specific cognitive probes inspired by therapeutic interventions and computational models of behavior may allow us to uncover the cognitive mechanisms targeted by common forms of psychotherapy. In turn, this may provide the basis for choosing effective psychotherapeutic interventions for given individuals.

In this study, using a randomised control trial design, we assess the effectiveness of a cognitive intervention aimed at repairing the potential breakdown of cooperation in the RTG from a low investment sent by a computerized investor. The intervention focuses on explaining the potential harm from reciprocating non-cooperative actions and suggesting a non-impulsive course of action to coax the investor back into cooperation. 
The computerized investor was programmed to play according to a hidden Markov model estimated from real players' data in prior research. A key aspect of this model is that the actions of the investor depend on a latent "trust state" which reacts dynamically to the trustees' returns simulating real-life trust-building scenarios. Such a model, informed by empirical data and integrating a responsive trust mechanism, represents a novel approach to study interactive behavior in multi-player games whilst keeping a high degree of experimental control. To foreshadow our results, we find that the intervention led to more cooperative actions (higher returns) by the participants and countered a tendency to send back lower returns after a transgression from the investor. However, we found no evidence that the effects of the intervention transfer to a different (Repeated Prisoner’s Dilemma) game.

@@

A total of 318 participants were recruited on the Prolific Academic platform (prolific.co). The mean age of participants was `r round(mean(as.numeric(d_finished$exitSurvey.age), na.rm=TRUE),1)` years, with a `r round(sd(as.numeric(d_finished$exitSurvey.age), na.rm=TRUE),1)` years standard deviation. Participants were paid a fixed fee of £5 plus a bonus payment dependent on their performance that averaged £0.71.

@@

The experiment had a 2 (Condition: Intervention or Control) by 3 (Game: Trust-Game Pre Intervention, Trust-Game Post Intervention, Prisoner's Dilemma Post Intervention) design, with repeated measures on the second factor. Participants were randomly assigned to one of the two levels of the first factor. The games were designed and implemented online using Empirica [@almaatouq2021empirica]. This research received approval from a UCL ethics board (ID:21029/001) and the experiment was performed in accordance with the ethics board guidelines and regulations.

@@

Participants played a 15-round Repeated Trust Game [@joyce_trust_1995] in the trustee role against a computer-programmed investor. On each round the investor is endowed with 20 units and decides how much of that endowment to invest. This investment is tripled and the trustee then decides how to split this tripled amount between them and the investor. If the trustee returns more than one third of the amount, the investor makes a gain. Each player was represented with an icon with the participant always on the left of the screen and the co-player on the right (Figure \ref{fig:HMMPanels}.a). The participants were able to choose the icon that represents them at the start of the experiment. The icon representing the co-player changed at the start of each new game, to simulate a new interaction partner. Participants were not told they were facing computerised co-players.

The strategy of the computerised investor was modelled on behavior of human investors in the Repeated Trust Game (RTG)  over 10-rounds with the same (human) co-player. Full detail on the datasets used in the Supplementary Information. Using this data, we estimated a hidden Markov model (HMM) on investors' behavior with three latent states. Each latent state was associated with a state-conditional distribution over the possible investments from 0 to 20 (Figure \ref{fig:HMMPanels}.b). These distributions reflect "low-trust", "medium-trust", or "high-trust". Over rounds, the investor can move between states, and the probability of these transitions was modelled as a function of their net return (i.e return - investment) in the previous round (see Figure \ref{fig:HMMPanels}.d). In order to instigate a potential breakdown of trust, thereby allowing us to probe efforts to repair trust, the computerised agent was programmed to provide a low investment on round 12 (pre-intervention) or round 13 (post-intervention). On all other rounds, the investor's actions were determined by randomly drawing an investment from the state-conditional distribution, with the state over rounds determined by randomly drawing the next state from the state-transition distribution as determined from the net return on the previous round (disregarding the net return immediately after the pre-programmed low investment rounds). The initial state for the HMM investor in each instance of the game was the "mid-trust" state.

On each round, immediately after being informed of the investment sent, participants in the intervention condition were asked to provide an evaluation of their emotion in terms of valence (from negative to positive) and arousal (from low to high). Participants in the control condition were asked to evaluate the investment in terms of speed (from slow to fast) and magnitude (from low to high). These evaluations were made by clicking on a two-dimensional field with labelled axes as shown in Figure \ref{fig:HMMPanels}.c for the intervention condition.

@@

At the start of the experiment, participants provided informed consent and were instructed the study would consist of three phases in which they would face a different other player. Participants were told their goal was to maximise the number of points in all phases. They were not told the number of rounds of each phase. Phase one was a 15 round Repeated Trust Game (RTG) in which participants took the role of trustee, facing the same investor over all 15 rounds. On each round, after being informed about the amount sent by the investor, participants were asked to evaluate their emotion (intervention condition) or the investment (control condition). Participants then decided on how much of the tripled investment to return to the investor, before continuing to the next round. After completing 15 rounds of the RTG, participants rated how cooperative, selfish, trustworthy and friendly they perceived the investor to be (all on a scale from 1 to 10). After phase one, participants in the intervention condition completed the intervention, and participants in the control condition solved anagrams. Subsequent phase two was similar to phase one, with participants being told they would face a new player. Phase three consisted of 7 rounds of the Repeated Prisoner's Dilemma game (RPD), with participants informed they would face a third player. Participants then completed questionnaires related to mentalising abilities, emotion regulation, and BPD traits (see the supplement for details). They were then asked about the strategy in the games, as well as whether they thought the other players were human or computer agents. They were then debriefed and thanked for their participation.

@@

The model was estimated using the `afex` package [@singmann_afex_2022] in R [@R_package]. More complex models with additional random effects could not be estimated reliably, and as such the estimated model can be considered to include the optimal random effects structure [@matuschek2017balancing]. A similar process was used to establish the random effects structures of other linear mixed-effects models used throughout the statistical analyses. There is no agreed upon way to calculate effect sizes for mixed effects models, as such we will report instead on testing differences in marginal means. For the $F$-tests, we used the Kenward-Roger approximation to the degrees of freedom, as implemented in the R package "afex". Unless otherwise stated, the default adjustment method for multiple comparison is Tukey's HSD. We Z-transform the Investment variable as centering is beneficial to interpreting the main effects more easily in the presence of interactions.

To model participants' returns in the RTG across games and conditions, we fit various hidden Markov models [@visser2022hidden] to participants' returns using the depmixS4 package [@visser_depmixs4_2021] for R. The transition between latent states is assumed to depend on the investment received and a dummy variable to characterise the group that the participant belongs to. Details on how the models are constructed can be found in the supplement. We fit models with different numbers of hidden states, and use the Bayesian Information Criterion [@schwarz_estimating_1978-1] to select the best model.

\begin{landscape}
```{r HMMPanels, fig.cap= "\\small{a: Screenshot of the repeated Trust Game. The game is played in dyads, with one player assigned the role of the investor and the other player that of the trustee. The investor is endowed with 20 units at the start of each round. They need to decide how much of that endowment they want to invest with the trustee. The investment is then multiplied by a factor of 3 and sent to the trustee who needs to decide how much of the multiplied investment they want to send back to the investor. Shown here is the stage at which the trustee decides how much to send back to the investor. Immediately before making this decision, participant are shown a two dimensional grid (Panel c) where they need to report their reaction to the investment. Those in the intervention condition were asked questions about their emotional state and those in the control condition were asked to report features of the investment unrelated to their emotion. Panels b - d: We construct the artificial investor agent by fitting a three-state hidden Markov model to data of human investors engaged in the 10 round Repeated Trust Game against human trustees. From the fitted HMM, we get the distribution of investments by the artificial investor agent conditional on its latent state as shown in Panel b. The fitted HMM also yields the transition probability of the agent to a state on trial t+1 as a function of the net return (difference between the investment sent and the amount received in return) on trial t as shown in Panel d. Each plot in Panel d represents a different starting latent state on trial t, and each line represents the probability of transitioning to a particular state in trial t+1.}",echo=FALSE, message=FALSE, warning=FALSE, out.width='1\\linewidth', out.height='1\\textheight', fig.align='center'}

@@

# Read in the images
image1 <- image_read("figures/trust_game.png")
image2 <- image_read("figures/HMMinvPolicy2.png")
image3 <- image_read("figures/emotion_grid.png")
image4 <- image_read("figures/HMMinvTrans2.png")

@@

# Annotate each image with a label
image1 <- image_annotate(image1, text = "a", location = "+200+10", size = 90, color = "black", gravity = "northwest")
image2 <- image_annotate(image2, text = "b", location = "+0+10", size = 90, color = "black", gravity = "northwest")
image3 <- image_annotate(image3, text = "c", location = "-0+10", size = 90, color = "black", gravity = "northwest")
image4 <- image_annotate(image4, text = "d", location = "+0+10", size = 90, color = "black", gravity = "northwest")

@@

# %>%
#   filter(complete.cases(.))   # <- one player had NA in one round as a return, so we exclude him from analysis.

@@

# average investment for first 10 rounds pre intervention
avg_pct <- avg_ret_df %>% filter(roundNum < 11 , gameNum.f == "pre") %>% summarise(avg_inv = mean(inv_pct), avg_ret = mean(ret_pct, na.rm=TRUE))
```

@@

To explore the HMM investors' behavior across games and conditions, we estimate a linear mixed-effects model of investments sent by the computerised HMM agent with Condition and Game and their interaction as fixed effects, and a similar random effects structure to the returns model. This shows a main effect of Condition, `r papaja::apa_print(mod_invs)$full_result$condition_f`<!--$F(1,317) = 8.7$, $p = .003$-->, and Game, `r papaja::apa_print(mod_invs)$full_result$gameNum_f`<!--$F(1,317) = 8.3$, $p = .004$-->. As can be seen in Figure \ref{fig:TrustGameResults}.d, investment was higher in the intervention compared to the control condition across games, and higher in the second game compared to first across conditions.

@@

>>>>

abstract: |
  Cooperation, underpinned by trust, is crucial for social interactions yet fragile once broken. This study investigates the effectiveness of a brief cognitive intervention in promoting resilient cooperative behavior in economic games. Using a randomized controlled online experiment, we examined how 318 participants, acting as trustees in the Repeated Trust Game (RTG), responded to an intervention inspired by Dialectical Behavior Therapy. Participants played two 15-round RTGs against a novel computer-simulated investor using a hidden Markov model (HMM) trained on human data, enabling adaptive, human-like behavior while maintaining experimental control. The intervention significantly increased cooperative behavior, with participants showing more resilient cooperation even after experiencing low investments. HMM analysis revealed that the intervention increased the probability of participants occupying high-cooperation states. However, these effects did not transfer to a subsequent Repeated Prisoner's Dilemma game. Our findings demonstrate the potential of targeted cognitive interventions to enhance cooperative behavior, while showcasing the value of HMM-based agents in studying dynamic social interactions.

@@

<!-- Cooperation refers to the collaborative efforts of individuals or entities working towards a shared goal or advantage [@tomasello2012two]. This coordinated approach often leads to outcomes that exceed the sum of individual contributions.  Trust is a key driver of cooperation, enabling individuals to engage in mutually beneficial interactions without fear of exploitation [@balliet2013trust]. This aligns with some definitions of trust as “a psychological state comprising the intention to accept vulnerability based on positive expectations of the intentions or behavior of another” [@rousseau1998not]. When we trust others, we are more likely to cooperate with them, even when there is a risk of being cheated. Prior research also emphasises the importance of social trust in determining why some people fare better than others physically and mentally [@giordano_trust_2016; @meng_multilevel_2014]. Research into the determinants of psychopathology has linked trust-based constructs to the emergence of mental health disorders [@fonagy_role_2014]. -->

Cooperation, defined as individuals or entities working together towards a shared goal, plays a fundamental role in promoting collective success and social harmony [@tomasello_two_2012]. At the heart of cooperation is trust—the belief that others will act in ways that are mutually beneficial, even when they have the opportunity to exploit the situation [@rousseau_introduction_1998]. Trust enables individuals to engage in risky interactions where immediate self-interest could easily override long-term benefits [@balliet_trust_2013]. Without trust, cooperation tends to break down, leading to suboptimal outcomes for all parties involved. Understanding how to maintain and repair cooperation following such breakdowns is therefore of significant interest to researchers and practitioners alike.

<!-- A well-established paradigm in the study of trust is the Repeated Trust Game (RTG; @joyce_trust_1995). In this game, the "investor" decides how much of an endowment to send to the other player (the "trustee"). The amount that is sent is tripled and the trustee decides, in return, how much of the tripled amount to send back to the investor. The Nash equilibrium for a single-round version of the game is for the investor to send nothing. In the repeated version, rewards for both players are maximised if they build trust and share the benefits of higher investments. If the investor is rewarded for taking the risk of sending an investment, they are likely invest again in future rounds. But if the investor obtains a low return on their investment, they will likely reduce future investments, thereby diminishing the potential gains for both players.  -->

The Repeated Trust Game (RTG) has emerged as a well-established paradigm for studying trust and cooperation in controlled settings [@joyce_trust_1995]. In this game, an "investor" decides how much of an endowment to send to a "trustee." The amount sent is typically multiplied by 3, and the trustee then decides how much of this multiplied amount to return to the investor. Cooperation emerges when both parties act in ways that promote mutual gains over multiple rounds. However, trust is fragile, and a single instance of defection—where one player fails to reciprocate appropriately—can lead to the breakdown of cooperation [@bendor_when_1991]. Once trust is violated, it is often difficult to re-establish, even if doing so would be mutually beneficial [@harth_spiral_2017].

Previous research has explored ways to encourage initial cooperation in trust games, such as using third-party enforcement [@charness_investment_2008; @fiedler_effect_2017], gratitude priming [@drazkowski_gratitude_2017], or with the concepts of friend and foe [@burnham_friend-or-foe_2000]. While these approaches improve early cooperative behavior, they often fail to address the more challenging task of repairing cooperation after trust has been broken. Once defection occurs, individuals may react impulsively by reducing their own cooperative efforts, even if re-establishing trust would be mutually beneficial. Thus, interventions aimed at restoring cooperation in these contexts are crucial but understudied.

<!-- Given the importance of trust for building and maintaining strong social bonds, interventions that foster and strengthen cooperation would be highly beneficial to society. Such interventions would allow people to more easily repair broken relationships and continue harvesting the benefits of cooperation even in the presence of accidental or intentional social norm violations. Previous studies have attempted to restore cooperation in trust games through mechanisms such as third-party enforcement [@charness_investment_2008; @fiedler_effect_2017] or gratitude priming [@drazkowski_gratitude_2017]. Whilst these results show that it is possible to improve cooperative outcomes at the start of the game, they do not address how to repair a breakdown of trust after intentional or accidental non-cooperative actions by the players. -->

<!-- To encourage the emergence and maintenance of trust in the RTG setting, some studies have focused on modification of the game, e.g. through introducing a third-party who monitors the actions of the other players [@charness_investment_2008; @fiedler_effect_2017]. Others chose to prime participants towards feeling gratitude [@drazkowski_gratitude_2017] or with the concepts of friend and foe [@burnham_friend-or-foe_2000]. Whilst these results show that it is possible to improve cooperative outcomes at the start of the game, they do not address how to repair a breakdown of trust after intentional or accidental non-cooperative actions by the players. Cooperation in the RTG can easily break down after a transgressive behavior such as a nil or very low investment by the investor or a return by the trustee below the investment sent [@bendor_when_1991]. Such breakdowns of cooperation are particularly evident when the trustee suffers from social disorders such as Borderline Personality Disorder (BPD; @lieb_borderline_2004). Trustees with BPD fail to engage in trust-repairing behaviors such as coaxing the investor by signalling trustworthiness via sending high returns [@king-casas_rupture_2008].  Neuroimaging analysis from the study of @king-casas_rupture_2008 is consistent with the idea that individuals with BPD may not recognize how offering low returns in such interactions conflicts with societal expectations. Further, computational analysis by @hula2018model links this issue to some BPD individuals' mentalizing difficulties, such as their inability to foresee the impact of their low returns on the investor's future actions. -->

In this study, we focus on the role of the trustee in the RTG, as a driver of cooperation dynamics. While the trustee does not exercise trust in the same way as the investor, their decisions to reciprocate (or not) play a critical role in maintaining or disrupting the cooperative relationship. A lack of reciprocation from trustees erodes trust over time [@servatka_words_2011] and is particularly evident when the trustee suffers from some personality disorders [@lieb_borderline_2004]. These trustees fail to engage in trust-repairing behaviors such as coaxing the investor by signalling trustworthiness via sending high returns [@king-casas_rupture_2008].
Additionally, unpredictable behavior from trustees fosters mistrust and impedes future cooperation [@rigdon_sustaining_2007]. In contrast, consistent cooperation from trustees promotes trust and encourages further collaboration, as evidenced by neural data [@king-casas_getting_2005]. Therefore, emphasizing the role of trustees in rebuilding cooperation is essential; when trustees demonstrate reliability and reciprocity, even after breaches of trust, cooperation can be restored and sustained.

Given the pivotal role of trustees' behavior in shaping cooperative dynamics, it is worth exploring whether interventions aimed at improving interpersonal skills could positively influence their decision-making in the RTG. In the broader field of psychological therapies, cognitive interventions inspired by  Dialectical Behavior Therapy [@linehan_cognitive-behavioral_1993] and Mentalisation Based Therapy [@allen_handbook_2006-1] have shown promise in enhancing social skills among individuals with interpersonal difficulties. These approaches often focus on helping individuals recognize the impact of their actions on others and consider alternative behavioral strategies. However, response to these treatments is highly variable, and determining which interventions are effective for particular sub-groups of patients is challenging [@rudge_mechanisms_2020; @arch_longitudinal_2012]. One promising approach is the study of how specific components of psychotherapeutic treatment affect quantitative markers of behavior such as those inferred through computational models [@huys_computational_2016; @reiter_neuro-cognitive_2021; @dercon_core_2024]. Combining the use of specific cognitive probes inspired by therapeutic interventions and computational models of behavior may allow us to uncover the cognitive mechanisms targeted by common forms of psychotherapy. In turn, this may provide the basis for choosing effective psychotherapeutic interventions for given individuals, potentially extending to improving trustee behavior and fostering cooperation in the RTG.

<!-- In devising interventions to repair trust in the RTG, we can take inspiration from cognitive interventions championed by successful therapies that aim to improve social dysfunction in BPD. Psychotherapies such as Mentalisation Based Therapy [@allen_handbook_2006-1] and Dialectical Behavior Therapy [@linehan_cognitive-behavioral_1993] have been shown to improve social skills in BPD [@gunderson_borderline_2018]. However, response to these treatments is highly variable, and determining which interventions are effective for particular sub-groups of patients is challenging [@rudge_mechanisms_2020; @arch_longitudinal_2012]. One promising approach is the study of how specific components of psychotherapeutic treatment affect quantitative markers of behavior such as those inferred through computational models [@huys_computational_2016; @reiter_neuro-cognitive_2021; @dercon2022core]. Combining the use of specific cognitive probes inspired by therapeutic interventions and computational models of behavior may allow us to uncover the cognitive mechanisms targeted by common forms of psychotherapy. In turn, this may provide the basis for choosing effective psychotherapeutic interventions for given individuals. -->

Drawing inspiration from such therapeutic approaches, we employ a randomized control trial to evaluate a cognitive intervention aimed at repairing cooperation after low investments from a computerized investor. The intervention in this study is a brief, multi-component cognitive intervention inspired by Dialectical Behavior Therapy (DBT) principles. It combines elements aimed at understanding long-term consequences of actions and promoting prosocial behavior. This approach mirrors real-world cognitive interventions that often employ multiple strategies to effect behavioral change [@linehan_dbt_2015]. Specifically, we hypothesize that encouraging participants to reflect on the consequences of their actions and consider a non-impulsive course of action might lead to more resilient cooperative behavior, even in the face of perceived non-cooperation from their partner.

We conducted an online experiment with 318 participants acting as trustees in two RTG rounds, with the intervention administered between rounds to a subset of participants. While previous studies have often relied on predetermined or simplistic computer strategies in economic games, our study introduces a novel approach using Hidden Markov Models (HMMs) to create more realistic, adaptive computer agents. A key aspect of these agents is that their actions depend on a latent “trust state” which reacts dynamically to the trustees’ returns, simulating real-life trust-building scenarios. To foreshadow our results, we find that the intervention led to more cooperative actions (higher returns) by the participants and countered a tendency to send back lower returns after a transgression from the investor. However, we found no evidence that the effects of the intervention transfer to a different (Repeated Prisoner’s Dilemma) game.

@@

A total of 320 participants were recruited on the Prolific Academic platform (prolific.co). Two players had incomplete trust game entries and their data was disregarded, leaving data for 318 participants for analysis, equally split between the two conditions. The required sample size was determined using an *a priori* power analysis to have an 80% probability to detect a small effect size (Cohen's f = 0.10) for a within-between interaction with a 5% type I error rate in a repeated measures ANOVA. The sample size calculation assumed 2 groups, 2 measurement per group and was performed using the G*Power software [@faul_statistical_2009]. The mean age of participants was `r round(mean(as.numeric(d_finished$exitSurvey.age), na.rm=TRUE),1)` years, with a `r round(sd(as.numeric(d_finished$exitSurvey.age), na.rm=TRUE),1)` years standard deviation. Participants were paid a fixed fee of £5 plus a bonus payment dependent on their performance that averaged £0.71.

@@

The experiment employed a 2 (Condition: Intervention or Control) by 2 (Game: Trust-Game Pre-Intervention, Trust-Game Post-Intervention) design, with repeated measures on the second factor. Participants were randomly assigned to either the intervention or control condition. Following the post-intervention Trust Game, all participants completed a Repeated Prisoner's Dilemma (RPD) game to assess potential transfer effects (See Figure \ref{fig:HMMPanels}.A for experiment overview). The games were designed and implemented online using Empirica [@almaatouq_empirica_2021-1]. This research received approval from a UCL ethics board (ID:21029/001) and the experiment was performed in accordance with the ethics board guidelines and regulations.

@@

Participants played two separate 15-round Repeated Trust Games [@joyce_trust_1995] in the role of trustee. They were told (correctly) that each 15-round game was played against the same co-player, and that they will face a new player when they started a new 15-round game. Participants were led to believe they were interacting with human co-players when in fact they faced the same computerised investor in all trust games. This design allows us to examine the effects of our intervention while controlling for individual differences in play style, as well as minimizing the influence of reputational concerns or carry-over effects that might occur if participants believed they were facing the same partner twice. In each round of the RTG, the investor (computer-simulated) was endowed with 20 units and decided how much of that endowment to invest. This investment was tripled, and the participant, as the trustee, then decided how to split this tripled amount between themselves and the investor. If the trustee returns more than one third of the amount, the investor makes a gain. On each round, immediately after being informed of the investment sent, participants in the intervention condition were asked to provide an evaluation of their emotion in terms of valence (from negative to positive) and arousal (from low to high). Participants in the control condition were asked to evaluate the investment in terms of speed (from slow to fast) and magnitude (from low to high). These evaluations were made by clicking on a two-dimensional field with labelled axes as shown in Figure \ref{fig:HMMPanels}.C for the intervention condition.

<!-- Participants played a 15-round Repeated Trust Game [@joyce_trust_1995] in the trustee role against a computer-programmed investor. On each round the investor is endowed with 20 units and decides how much of that endowment to invest. This investment is tripled and the trustee then decides how to split this tripled amount between them and the investor. If the trustee returns more than one third of the amount, the investor makes a gain. Each player was represented with an icon with the participant always on the left of the screen and the co-player on the right (Figure \ref{fig:HMMPanels}.A). The participants were able to choose the icon that represents them at the start of the experiment. The icon representing the co-player changed at the start of each new game, to simulate a new interaction partner. Participants were not told they were facing computerised co-players. -->

### Adaptive Computerised Investor {.unnumbered}

The strategy of the computerised investor was modelled on behavior of human investors in the Repeated Trust Game (RTG) over 10-rounds with the same (human) co-player from existing datasets. Full detail on the data sources used are in the Supplementary Information. Using this data, we estimated a hidden Markov model (HMM) on investors' behavior with three latent states. Each latent state was associated with a state-conditional distribution over the possible investments from 0 to 20 (Figure \ref{fig:HMMPanels}.B). These distributions reflect "low-trust", "medium-trust", or "high-trust". Over rounds, the investor can move between states, and the probability of these transitions was modelled as a function of their net return (i.e return - investment) in the previous round (see Figure \ref{fig:HMMPanels}.D). In order to instigate a potential breakdown of trust, thereby allowing us to probe efforts to repair trust, the computerised agent was programmed to provide a low investment on round 12 (pre-intervention) or round 13 (post-intervention). On all other rounds, the investor's actions were determined by randomly drawing an investment from the state-conditional distribution, with the state over rounds determined by randomly drawing the next state from the state-transition distribution as determined from the net return on the previous round (disregarding the net return immediately after the pre-programmed low investment rounds). The initial state for the HMM investor in each instance of the game was the "mid-trust" state.

Our use of HMMs to model investor behavior represents a significant advancement in experimental design for economic games. Unlike traditional fixed-strategy computer opponents, our HMM-based agents adapt their behavior based on the participant's actions, mirroring the complex decision-making processes observed in human players. This approach allows us to maintain experimental control while significantly enhancing the ecological validity of our study.

@@

At the start of the experiment, participants provided informed consent and were instructed the study would consist of three phases. Participants across conditions were told their goal was to maximise the number of points in all phases. Participants had to pass comprehension checks about the number of phases, the fact the co-player was the same within each phase, and that they would face a new player at the start of each new phase. They were not told the number of rounds of each phase.

Phase one was a 15 round RTG in which participants took the role of trustee, facing the same investor over all 15 rounds. Participants were given details instructions about the game and had to pass comprehension checks to test their understanding of their and their co-player's payoff in a hypothetical situation. On each round, after being informed about the amount sent by the investor, participants were asked to evaluate their emotion (intervention condition) or the investment (control condition). Participants then decided on how much of the tripled investment to return to the investor, before continuing to the next round. After completing 15 rounds of the RTG, participants rated how cooperative, selfish, trustworthy and friendly they perceived the investor to be (all on a scale from 1 to 10). After phase one, participants in the intervention condition completed the intervention, and participants in the control condition solved anagrams. Subsequent phase two was similar to phase one, with participants being told they would face a new player.

Phase three consisted of 7 rounds of the Repeated Prisoner's Dilemma game (RPD), with participants informed they would face a third player. Participants then completed questionnaires related to mentalising abilities, emotion regulation, and BPD traits (see the supplement for details). They were then asked about the strategy in the games, as well as whether they thought the other players were human or computer agents. Finally, participants were debriefed and thanked for their participation.

@@

Our choice of linear mixed-effects models (LMMs) over mixed ANOVA was based on their greater flexibility in handling our complex data structure, including continuous predictors and nested repeated measures. LMMs offer increased statistical power and more flexible assumptions, particularly regarding sphericity, which is often violated in repeated measures designs. The model was estimated using the `afex` package [@singmann_afex_2022] in R. More complex models with additional random effects could not be estimated reliably, and as such the estimated model can be considered to include the optimal random effects structure [@matuschek_balancing_2017]. A similar process was used to establish the random effects structures of other linear mixed-effects models used throughout the statistical analyses. As there is no agreed upon way to calculate effect sizes for mixed effects models, we will report instead on testing differences in marginal means. For the $F$-tests, we used the Kenward-Roger approximation to the degrees of freedom, as implemented in the R package "afex". For all post-hoc pairwise comparisons following significant effects in the mixed-effects models, we used Tukey's Honestly Significant Difference (HSD) test to adjust for multiple comparisons, unless otherwise stated. We Z-transform the Investment variable as centering is beneficial to interpreting the main effects more easily in the presence of interactions.

To model participants' returns in the RTG across games and conditions, we fit various hidden Markov models [@visser_hidden_2022] to participants' returns using the depmixS4 package [@visser_depmixs4_2021] for R. The transition between latent states is assumed to depend on the investment received and a dummy variable to characterise the group that the participant belongs to. Details on how the models are constructed can be found in the supplement. We fit models with different numbers of hidden states, and use the Bayesian Information Criterion [@schwarz_estimating_1978-1] to select the best fitting model.

\begin{landscape}
```{r HMMPanels, fig.cap= "\\small{Panel A: Experimental design overview. Participants in both conditions played two 15-round Repeated Trust Games (RTGs) with human-like Hidden Markov Model (HMM) investors, separated by a manipulation phase. The intervention condition received a cognitive intervention, while the control condition solved anagrams. Both groups then completed a 7-round Repeated Prisoner's Dilemma (RPD) with a tit-for-tat (TFT) co-player to assess transfer effects. This design allows for comparison of cooperative behavior before and after the intervention, as well as between conditions. Immediately before making return decisions in the RTG, participants are shown a two dimensional grid (Panel C) where they need to report their reaction to the investment. Those in the intervention condition were asked questions about their emotional state and those in the control condition were asked to report features of the investment unrelated to their emotion. Panels B - D: We construct the artificial investor agent by fitting a three-state hidden Markov model to data of human investors engaged in the 10 round Repeated Trust Game against human trustees. From the fitted HMM, we get the distribution of investments by the artificial investor agent conditional on its latent state as shown in Panel B. The fitted HMM also yields the transition probability of the agent to a state on trial t+1 as a function of the net return (difference between the investment sent and the amount received in return) on trial t as shown in Panel D. Each plot in Panel D represents a different starting latent state on trial t, and each line represents the probability of transitioning to a particular state in trial t+1.}",echo=FALSE, message=FALSE, warning=FALSE, out.width='1\\linewidth', out.height='1\\textheight', fig.align='center'}

@@

# Read in the images
image1 <- image_read("figures/timeline.png")
image2 <- image_read("figures/HMMinvPolicy2.png")
image3 <- image_read("figures/emotion_grid.png")
image4 <- image_read("figures/HMMinvTrans2.png")

@@

# Annotate each image with a label
image1 <- image_annotate(image1, text = "A", location = "+0+10", size = 90, color = "black", gravity = "northwest")
image2 <- image_annotate(image2, text = "B", location = "+0+10", size = 90, color = "black", gravity = "northwest")
image3 <- image_annotate(image3, text = "C", location = "-0+10", size = 90, color = "black", gravity = "northwest")
image4 <- image_annotate(image4, text = "D", location = "+0+10", size = 90, color = "black", gravity = "northwest")

@@

# average investment for first 10 rounds pre intervention
avg_pct <- avg_ret_df %>% filter(roundNum < 11 , gameNum.f == "pre") %>% summarise(avg_inv = mean(inv_pct), avg_ret = mean(ret_pct, na.rm=TRUE))

# Find rows with NAs, excluding the 'ret_pct' column
df_temp <- avg_ret_df%>%dplyr::select(playerId,condition.f,investment,returns,roundNum)

# Identify the playerId(s) with NA values
na_player <- df_temp[!complete.cases(df_temp), "playerId"]

# Remove rows corresponding to the playerId with NA values
avg_ret_df <- avg_ret_df[!avg_ret_df$playerId %in% na_player, ]

# Display the cleaned dataframe
avg_ret_df

@@

@@

To explore the HMM investors' behavior across games and conditions, we estimate a linear mixed-effects model of investments sent by the computerised HMM agent with Condition and Game and their interaction as fixed effects, and a similar random effects structure to the returns model. This shows a main effect of Condition, `r papaja::apa_print(mod_invs)$full_result$condition_f`<!--$F(1,317) = 8.7$, $p = .003$-->, and Game, `r papaja::apa_print(mod_invs)$full_result$gameNum_f`<!--$F(1,317) = 8.3$, $p = .004$-->. As can be seen in Figure \ref{fig:TrustGameResults}.D, investment was higher in the intervention compared to the control condition across games, and higher in the second game compared to first across conditions.

@@

# Filter the dataframe for PBOR scores higher than 40 and group by condition
result <- avg_ret_df %>%
  filter(PBOR_score >= 40) %>%
  group_by(condition.f) %>%
  summarise(count = n())
